---
title: "Benchmarking DelayedArray branch of **minfi**"
author: "Peter Hickey"
date: "`r BiocStyle::doc_date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
editor_options: 
  chunk_output_type: console
---

## Setup

```{r, include = FALSE, setup}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", collapse = TRUE,
                      message = TRUE, warning = TRUE)
```

```{r}
suppressPackageStartupMessages(library(minfi))
suppressPackageStartupMessages(library(minfiData))
```

## Using **minfi** v1.25.1

```{r 1.25.1}
RGSet <- minfiData::RGsetEx
system.time(MSetRaw_v1.25.1 <- minfi::preprocessRaw(RGSet))
system.time(MSetSWAN_v1.25.1 <- minfi::preprocessSWAN(RGSet))
system.time(MSetIllumina_v1.25.1 <- minfi::detectionP(RGSet))
system.time(detP_v1.25.1 <- minfi::detectionP(RGSet))

MSet <- minfiData::MsetEx
system.time(MSetFixedOutliers_v1.25.1 <- minfi::fixMethOutliers(MSet))
```

## Using DelayedArray branch of **minfi**

Some setting up:

```{r}
suppressPackageStartupMessages(library(DelayedArray))
suppressPackageStartupMessages(library(HDF5Array))
suppressPackageStartupMessages(library(DelayedMatrixStats))

source("R/utils.R")
source("R/DelayedArray_utils.R")
source("R/preprocessRaw.R")
source("R/preprocessIllumina.R")
source("R/detectionP.R")
source("R/minfiQC.R")
source("R/preprocessSwan.R")

DelayedArray:::set_verbose_block_processing(TRUE)

DEFAULT_BLOCK_SIZE <- getOption("DelayedArray.block.size")
DEFAULT_BLOCK_SIZE
```

Create the various _RGChannelSet_ and _MethylSet_ objects:

```{r}
RGSet_in_memory_DelayedMatrix <- minfiData::RGsetEx
system.time(
    assays(RGSet_in_memory_DelayedMatrix) <- endoapply(
        assays(RGSet_in_memory_DelayedMatrix),
        DelayedArray)
)

RGSet_on_disk_DelayedMatrix <- minfiData::RGsetEx
system.time(
    assays(RGSet_on_disk_DelayedMatrix) <- endoapply(
        assays(RGSet_on_disk_DelayedMatrix),
        writeHDF5Array))

MSet_in_memory_DelayedMatrix <- minfiData::MsetEx
system.time(
    assays(MSet_in_memory_DelayedMatrix) <- endoapply(
        assays(MSet_in_memory_DelayedMatrix),
        DelayedArray)
)

MSet_on_disk_DelayedMatrix <- minfiData::MsetEx
system.time(
    assays(MSet_on_disk_DelayedMatrix) <- endoapply(
        assays(MSet_on_disk_DelayedMatrix),
        writeHDF5Array))
```

### Using an ordinary _matrix_-backed _RGChannelSet_

```{r}
system.time(MSetRaw <- preprocessRaw(RGSet))
system.time(MSetIllumina <- preprocessIllumina(RGSet))
system.time(MSetSWAN <- preprocessSWAN(RGSet))
system.time(detP <- detectionP(RGSet))

system.time(MSetFixedOutliers <- fixMethOutliers(MSet))
```

### Using an in-memory _DelayedMatrix_-backed _RGChannelSet_

#### Default block size

```{r}
options(DelayedArray.block.size = DEFAULT_BLOCK_SIZE)
system.time(
    MSetRaw_in_memory_DelayedMatrix <- preprocessRaw(RGSet_in_memory_DelayedMatrix))
system.time(MSetSWAN_in_memory_DelayedMatrix <- preprocessSWAN(
    RGSet_in_memory_DelayedMatrix))
system.time(
    MSetIllumina_in_memory_DelayedMatrix <- preprocessIllumina(
        RGSet_in_memory_DelayedMatrix))
system.time(
    detP_in_memory_DelayedMatrix <- detectionP(RGSet_in_memory_DelayedMatrix))

system.time(
    fixMethOutliers_in_memory_DelayedMatrix <- fixMethOutliers(
        MSet_in_memory_DelayedMatrix))

```

#### Increased block size

```{r}
options(DelayedArray.block.size = DEFAULT_BLOCK_SIZE * 100L)
system.time(MSetRaw_in_memory_DelayedMatrix <- preprocessRaw(
    RGSet_in_memory_DelayedMatrix))
system.time(
    MSetSWAN_in_memory_DelayedMatrix <- preprocessSWAN(
        RGSet_in_memory_DelayedMatrix))
system.time(
    MSetIllumina_in_memory_DelayedMatrix <- preprocessIllumina(
        RGSet_in_memory_DelayedMatrix))
system.time(
    detP_in_memory_DelayedMatrix <- detectionP(RGSet_in_memory_DelayedMatrix))

system.time(
    fixMethOutliers_in_memory_DelayedMatrix <- fixMethOutliers(
        MSet_in_memory_DelayedMatrix))
```

### Using an on-disk (HDF5) _DelayedMatrix_-backed _RGChannelSet_

Using the **HDF5Array** backend:

```{r}
setRealizationBackend("HDF5Array")
```

#### Default block size

```{r}
options(DelayedArray.block.size = DEFAULT_BLOCK_SIZE)
system.time(
    MSetRaw_on_disk_DelayedMatrix <- preprocessRaw(RGSet_on_disk_DelayedMatrix))
system.time(
    MSetSWAN_on_disk_DelayedMatrix <- preprocessSWAN(
        RGSet_on_disk_DelayedMatrix))

system.time(
    MSetIllumina_on_disk_DelayedMatrix <- preprocessIllumina(
        RGSet_on_disk_DelayedMatrix))
system.time(
    detP_on_disk_DelayedMatrix <- detectionP(RGSet_on_disk_DelayedMatrix))

system.time(
    fixMethOutliers_in_memory_DelayedMatrix <- fixMethOutliers(
        MSet_on_disk_DelayedMatrix))
```

#### Increased block size

```{r}
options(DelayedArray.block.size = DEFAULT_BLOCK_SIZE * 100L)
system.time(
    MSetRaw_on_disk_DelayedMatrix <- preprocessRaw(RGSet_on_disk_DelayedMatrix))
system.time(
    MSetSWAN_on_disk_DelayedMatrix <- preprocessSWAN(
        RGSet_on_disk_DelayedMatrix))
system.time(
    MSetIllumina_on_disk_DelayedMatrix <- preprocessIllumina(
        RGSet_on_disk_DelayedMatrix))
system.time(
    detP_on_disk_DelayedMatrix <- detectionP(RGSet_on_disk_DelayedMatrix))

system.time(fixMethOutliers_in_memory_DelayedMatrix <-
                fixMethOutliers(MSet_on_disk_DelayedMatrix))
```
